---
title: "TruMedia Hackathon 2017"
author: "Christopher Teixeira"
output: 
  html_document:
    theme: flatly
    keep_md: true
---

This document represents the work of Christopher Teixeira for the 2017 TruMedia Hackathon. 

# Understanding pitches 
For illustrative purposes, I wanted to understand this new field _probCalledStrike_ to see how well it might be used to determine a pitcher's best pitch. 

```{r readindata,echo=F,warning=F,cache=T}
library(doMC)
registerDoMC(cores = 3)

library(readr)
library(dplyr)
library(lubridate)
d <- read_csv("../../2016.csv")

# Get a glimpse of the data
glimpse(d)

# Create additional features
d <- d %>%
  mutate(battingTeam = ifelse(side=="T", visitor, home),
         date = date(ymd_hms(gameDate)))

# Create a pitchID for ordering pitches
d <- d %>% 
  group_by(gameString,pitcherId) %>%
  mutate(pitchID = row_number()) %>%
  ungroup()
```

```{r bestPitch, cache=T}
bp <- d %>%
  filter(pitcher=="Jon Lester",
         pitchType!="PO",
         pitchType!="UN") %>%
  group_by(pitcher, pitchType) %>%
  summarise(numPitches= n(),
            probCalledStrikeMean = mean(probCalledStrike),
            probCalledStrikeSd = sd(probCalledStrike)) %>%
  arrange(desc(probCalledStrikeMean))

knitr::kable(bp)

library(ggplot2)
ggplot(data=d %>% filter(pitcher=="Jon Lester",pitchType!="PO",pitchType!="UN"),
       aes(x=pitchType,y=probCalledStrike)) +
  geom_boxplot() +
  geom_jitter(width = 0.2)
```

We can see that for Jon Lester, he throws the cutter the most, and gets a higher average _probCalledStrike_. This matches up against [Brooks Baseball's Jon Lester card](http://www.brooksbaseball.net/landing.php?player=452657) which provides a nice validation for this metric.

My next question was, "could you use _probCalledStrike_ as a way to determine whether he will throw that type of pitch again?" This gets into a bit of a sophisticated solution. Using previous pitch information, we can try and see if it helps to determine whether he feels confident in throwing that pitch. 

In starting to look at this, I'm taking a very few amount of features. Here's a quick description of the features.

```{r features, cache=T}
library(dplyr)

df <- d %>% 
  filter(pitchType!="UN",
         pitchType!="PO",
         pitcher=="Jon Lester"
         ) %>%
  select(pitcher,pitchID,pitchType,probCalledStrike,pitchResult,batterHand,umpireId,
         balls, strikes, outs, manOnFirst, manOnSecond, manOnThird, inning, timesFaced) %>%
  mutate(lastPitchType=lag(pitchType),
         lastprobCallStrike=lag(probCalledStrike),
         lastPitchResult = lag(pitchResult),
         lastBatterHand = lag(batterHand)) %>%
  filter(pitchID>1) %>%
  select(-pitchResult,-pitchID,-probCalledStrike,-pitcher) 

df$pitchType <- factor(df$pitchType)
df$lastPitchType <- factor(df$lastPitchType)
df$lastPitchResult <- factor(df$lastPitchResult)
df$batterHand <- factor(df$batterHand)
df$lastBatterHand <- factor(df$lastBatterHand)
df$umpireId <- factor(df$umpireId)

knitr::kable(data.frame(Features=names(df),
                        Description=c("The pitch type being predicted",
                                      "The batter's hand",
                                      "The umpire making the call",
                                      "The number of balls for the at bat before the pitch",
                                      "The number of strikes for the at bat before the pitch",
                                      "The number of outs before the pitch",
                                      "Boolean for a runner on first",
                                      "Boolean for a runner on second",
                                      "Boolean for a runner on third",
                                      "Inning",
                                      "Number of times batter faced this pitcher within this game",
                                      "The last pitch type thrown",
                                      "The last pitch probability of called strike",
                                      "The last pitch result",
                                      "The batter's hand against the last pitch")))
```

Now that we have a set of features to work with, let's get into modeling. For this exercise, I chose a random forest for its ability to predict multiple classes and speed in execution. I chose to use the [caret](http://topepo.github.io/caret/index.html) library for the ability to switch models later on just in case it is needed. 

First up, let's split the data into train and test data sets. This will allow us better control to measure the performance of the model. I'll also take the time to create the train control that will be used in the train function. 

```{r datasplit, cache=T}
library(caret)

split.data <- createDataPartition(y=df$pitchType, 
                                  p = 0.6, 
                                  list=F)

df.train <- df[split.data,]
df.test <- df[-split.data,]
control <- trainControl(method="repeatedcv", 
                        number=10, 
                        repeats=3, 
                        savePredictions="final",
                        index=createResample(df.train$pitchType, 10),
                        classProbs=TRUE)
```

Now let's actually train the random forest and print out the model information. As you can see, this isn't exactly a great model. 

```{r randomforest, cache=T}
model <- train(pitchType~., data=df.train, method="multinom", trControl=control)
print(model)

confusionMatrix(model)

model.test <- predict(model, newdata = df.test,type="prob")
model.test$pred <- predict(model, newdata = df.test)
model.test$obs <- df.test$pitchType
confusionMatrix(model.test$pred,model.test$obs)
```

Well, let's try a different approach. By ensembling multiple models together, we can see if there's any value in taking the best from each model. 

```{r ensemble, cache=T}
library(caretEnsemble)

models <- caretList(pitchType~., 
                    data=df.train,
                    trControl=control,
                    methodList=c("rf", "nnet", "svmRadial")
                    )

results <- resamples(models)

summary(results)
modelCor(results)
dotplot(results)
splom(results)
```